\section{Literature Review}

An extensive literature review was conducted to examine existing research related to bird
species classification, fine-grained visual recognition, and citizen science systems.
Previous studies on bird species identification from images were analyzed in terms of
their machine learning architectures, dataset preparation strategies, evaluation metrics,
and overall model performance. Particular attention was given to recent studies that
clearly explain their model selection process, data acquisition methods, and experimental
results.

Early bird species recognition systems commonly relied on traditional machine learning
algorithms such as Support Vector Machines (SVM) and Random Forest classifiers.
These approaches generally used handcrafted features including color histograms,
shape descriptors, and texture patterns. However, such methods often struggle when
applied to real-world image datasets that contain variations in lighting conditions,
background complexity, image quality, and bird pose \cite{kambali2024automated,
lolge2025evolution}. Recent studies show that Convolutional Neural Networks (CNNs)
significantly outperform traditional machine learning approaches because they can
automatically learn hierarchical visual features directly from images
\cite{kambali2024automated}.

Bird species classification is commonly considered a Fine-Grained Visual Classification
(FGVC) problem because different bird species may share highly similar visual
characteristics. In FGVC tasks, models must distinguish subtle local differences such as
feather patterns, beak structure, wing shape, and color variations. Chen et al. emphasize
that successful FGVC systems require models capable of learning discriminative local
features and fine visual details \cite{chen2024fetfgvc}. Modern deep learning architectures,
particularly CNN- and Transformer-based models, provide the performance necessary for
handling these complex classification tasks.

Several studies compare the effectiveness of different CNN architectures for bird species
classification. Bilgin et al. evaluated architectures including VGG16, ResNet50, and
DenseNet and reported that DenseNet and ResNet achieved better classification
performance than VGG16 due to improved feature propagation and gradient flow
\cite{bilgin2022bird}. Similarly, Tang compared VGG, ResNet, InceptionV3, and
MobileNet architectures and found that ResNet and Inception-based models achieved
higher F1-scores, while MobileNet provided better computational efficiency
\cite{tang2025comparative}. These studies indicate that deeper architectures generally
achieve higher accuracy, whereas lightweight architectures may be more suitable for
deployment in real-world applications with limited computational resources.

Recent research increasingly emphasizes not only classification accuracy but also
inference speed, computational cost, and deployability
\cite{lolge2025evolution,tang2025comparative}. Lightweight architectures such as
MobileNet are considered advantageous for mobile and web-based applications due to
their lower hardware requirements and faster inference times. On the other hand,
architectures such as DenseNet and ResNet often achieve superior classification
performance at the expense of increased computational complexity
\cite{bilgin2022bird}. These findings are particularly important for this project because
the developed system aims to integrate machine learning inference into a practical citizen
science platform.

In addition to classification models, studies related to bird image datasets and citizen
science systems were also examined. Public biodiversity databases and citizen science
platforms provide large amounts of bird observation data that are valuable for training
machine learning models. One important example is the Global Biodiversity Information
Facility (GBIF), which offers openly accessible biodiversity records collected from
multiple sources worldwide \cite{gbif2025}. Such platforms contribute significantly to
large-scale ecological monitoring and dataset generation.

Citizen science has transformed biodiversity monitoring by enabling decentralized data
collection from large communities of volunteers. Fraisl et al. discuss how citizen science
has become an important component of environmental and ecological research due to its
ability to collect large-scale observational data \cite{fraisl2022citizen}. Similarly,
Bonney et al. explain that citizen science projects can improve public engagement and
scientific awareness while contributing valuable environmental data
\cite{bonney2015citizen}. However, citizen science data may also introduce challenges
including observer bias, class imbalance, inconsistent image quality, and incorrect
species labels \cite{johnston2022outstanding}. Baker et al. mention that automated
verification and validation techniques are increasingly being used to improve the quality
and reliability of citizen science datasets \cite{baker2021verification}.

The findings obtained from the literature review played an important role in guiding the
design and implementation decisions of this project. Existing studies helped identify
suitable deep learning architectures, appropriate evaluation metrics, and effective data
processing strategies. Furthermore, the literature review highlighted several common
challenges in bird species classification systems, including class imbalance, fine-grained
species similarity, background noise, and variation in image quality. Understanding these
challenges was important for designing a robust bird classification and citizen science
platform.

\clearpage